Negative Perceptions of Outsourcing to Artificial Intelligence

### 2025-04-28 manuscript.qmd
Start preparing supplementary materials
First draft of manuscript

---
title: "Negative Perceptions of Outsourcing to Artificial Intelligence"
shorttitle: "Outsourcing to AI"
author:
  - name: Scott Claessens
    corresponding: false
    orcid: 0000-0002-3562-6981
    email: scott.claessens@gmail.com
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
  - name: Pierce Veitch
    corresponding: false
    orcid: 0009-0005-3364-7470
    email: pv201@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
  - name: Jim Everett
    corresponding: true
    orcid: 0000-0003-2801-5426
    email: j.a.c.everett@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
abstract: "xxxx"
keywords: [keyword1, keyword2, keyword3]
bibliography: bibliography.bib
engine: knitr
format:
  apaquarto-pdf: default
execute:
  echo: false
  warning: false
  error: false
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Supplementary Figure
      space-before-numbering: true
      latex-list-of-description: Supplementary Figure
      caption-location: bottom
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Supplementary Table
      space-before-numbering: true
      latex-list-of-description: Supplementary Table
      caption-location: top
---

@@

@@

@@

@@

>>>>

---
title: "Negative Perceptions of Outsourcing to Artificial Intelligence"
shorttitle: "Outsourcing to AI"
author:
  - name: Scott Claessens
    corresponding: false
    orcid: 0000-0002-3562-6981
    email: scott.claessens@gmail.com
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - data curation
      - formal analysis
      - investigation
      - methodology
      - visualization
      - writing
      - editing
  - name: Pierce Veitch
    corresponding: false
    orcid: 0009-0005-3364-7470
    email: pv201@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - formal analysis
      - methodology
  - name: Jim Everett
    corresponding: true
    orcid: 0000-0003-2801-5426
    email: j.a.c.everett@kent.ac.uk
    affiliations:
      - name: University of Kent
        department: School of Psychology
        address: Keynes College
        city: Canterbury
        country: UK
        postal-code: CT2 7NP
    role:
      - conceptualization
      - funding acquisition
      - methodology
      - supervision
      - editing
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    study-registration: ~
    data-sharing: All data and original code can be found on GitHub at <https://github.com/ScottClaessens/outsourcing>.
    related-report: ~
    conflict-of-interest: The authors have no conflicts of interest to disclose.
    financial-support: This work was supported by ...
    gratitude: ~
    authorship-agreements: ~
abstract: >
  As artificial intelligence (AI) tools become increasingly integrated into 
  daily life, people are beginning to outsource not only professional tasks but
  also socio-relational ones. Large language models like ChatGPT can generate
  wedding vows, speeches, and personal messages, raising questions about how
  individuals who use AI for such tasks are perceived by others. In this paper,
  we conduct five pre-registered studies with British participants (*n* = 3649)
  to understand how people view those who outsource tasks to AI, and how this
  depends on how socio-relational the task is, whether AI is used as a tool or
  fully delegated to, and the acknowledgment of the AI use. We find negative
  perceptions of outsourcing, particularly for socio-relational tasks, and
  demonstrate that this does not only make us think more negatively about the
  outsourcer, but the outsourced work itself. We then turn to examine potential
  mechanisms underlying negative perceptions of outsourcing, investigating the
  roles of effort, perceived authenticity, and the reasons for using AI. Our
  research highlights the way that AI use shapes our perceptions of people,
  raising key philosophical questions about praise, blame, and authenticity in a
  world filled with AI-mediated interactions.
keywords: [artificial intelligence, character evaluation, outsourcing, trust]
bibliography: bibliography.bib
engine: knitr
format:
  apaquarto-pdf: default
execute:
  echo: false
  warning: false
  error: false
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Supplementary Figure
      space-before-numbering: true
      latex-list-of-description: Supplementary Figure
      caption-location: bottom
    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Supplementary Table
      space-before-numbering: true
      latex-list-of-description: Supplementary Table
      caption-location: top
floatsintext: true
---

@@

In recent years, artificial intelligence (AI) technologies have emerged as 
useful tools to save time and effort across a wide variety of tasks. Large 
language models, such as ChatGPT, Copilot, and DeepSeek, have been shown to 
increase efficiency for professional tasks like writing [@Noy2023], project 
planning [@Barcaui2023], and programming [@Peng2023]. In a 2022 Pew survey of
the American public, time-saving and efficiency were listed among the top
reasons for excitement about the increased presence of AI in daily life
[@Rainie2022].

As outsourcing routine tasks to AI becomes more commonplace, people may begin to
use AI tools to efficiently complete more socio-relational tasks. Consider, for
example, writing wedding vows. While wedding vows are widely understood to be an
opportunity for couples to express their love in their own words, large language
models like ChatGPT will dutifully produce wedding vows upon request, without
necessarily requiring any information about the partner in question. In
characteristic ChatGPT style, the resulting text will likely be generic enough
to apply to anyone, yet well-written enough to read aloud at the altar. A recent
Reddit post from a disgruntled newlywed suggests that AI-outsourcing of this
kind is already happening [@miramar0].

Of course, there is nothing new about outsourcing socio-relational tasks.
Ghostwriters have existed for years: in the 19th century, Victor Hugo, the
writer of *Les Misérables*, used to write love letters for others to earn extra
money [ref]. Today, the internet is abound with professional paid services for
writing wedding vows and personal speeches. But it remains unclear how people
who outsource tasks like these are perceived by others, particularly when the
tasks are outsourced to impersonal AI tools like ChatGPT.

There are several reasons to expect that people will be perceived negatively for
outsourcing tasks to AI. First, the effort put into a task is often used as a
signal of one's moral character [@Celniker2023; @Tissot2025]. People who free
ride on the efforts of others are perceived as less moral [@Cubitt2011], and
people's outputs are perceived as more valuable the more effort was ostensibly
put into them [@Kruger2004]. Second, people are given less credit when their
work was generated jointly with or entirely by AI instead of by themselves
[@Earp2024]. Third, from a virtue ethics perspective, people who outsource moral
tasks to AI skip the opportunity to cultivate virtuous habits and develop their
moral character [@Vallor2015]. These disparate lines of reasoning predict that
people will be perceived more negatively if they outsource tasks to AI, though
this has yet to be tested.

If people are perceived more negatively for outsourcing tasks to AI, there is 
also the question of how these negative perceptions might vary across different
types of tasks and different kinds of AI use. Will outsourcing wedding vows to
AI be perceived just as negatively as outsourcing a cover letter or a few lines
of computer code? Given that effort is thought to be a particularly important
cue for cooperative partner choice in social situations [@Celniker2023], we 
predict that outsourcing socio-relational tasks to AI will be perceived more 
negatively than routine or professional tasks.

It is also possible for people to outsource to AI in different ways. One can 
complete tasks by using AI as a collaborative tool, sculpting its answer over 
successive prompts and editing the resulting output by hand. Alternatively, one
can simply copy the AI's output verbatim. To the extent that the latter approach
requires less effort, we predict that "fully" outsourcing to AI will be
perceived more negatively. People can also differ in whether they are honest
about their use of AI or, in the words of the disgruntled newlywed, deceptively
"pass [the work] off as [their] own" [@miramar0]. Since people tend to dislike
and distrust those who are deceptive [@Tyler2006], it follows that people will
perceive deceptive AI-outsourcing more negatively.

Here, we test these predictions across five pre-registered experiments with 
British participants. In our initial pilot studies, we found that people who 
outsource a range of tasks to AI are perceived more negatively than people who
complete the tasks by themselves (see Supplementary Materials). In Study 1, we
expand on this by manipulating the type of outsourcing to AI (as a collaborative
tool vs. fully outsourcing) and whether people are honest or deceptive in their
usage of AI. In Study 2, we measure perceptions of both outsourcers and the
outsourced work. In Studies 3-5, we test potential mechanisms underlying
negative perceptions of outsourcing, including perceived effort, perceived
authenticity, and the reasons for using AI.

@@

## Methods

### Ethical Approval

Ethical approval was granted for all studies in this paper by the University of
Kent Psychology Research Ethics Panel. Participants in all studies provided
informed consent and were debriefed after the study.

### Participants

We conducted a power simulation to determine our target sample size. The 
simulation suggested that a sample size of 150 participants per condition
(overall *n* = 750 for five conditions) would be required to detect a small 
difference between conditions (Cohen's *d* ≈ 0.20) with above 80% power.

```{r}
study2_sample <-
  tar_read(study2_data) %>%
  group_by(id) %>%
  summarise(
    gender = unique(gender),
    age = unique(age),
    chatgpt_used = unique(chatgpt_used)
    )
```

We recruited a convenience sample of 800 participants from the United Kingdom 
through the online platform Prolific (<https://www.prolific.com/>). After 
excluding participants who failed our pre-treatment attention check, we were 
left with a final sample of `r nrow(study2_sample)` participants 
(`r sum(study2_sample$gender == "Female")` female; 
`r sum(study2_sample$gender == "Male")` male;
`r sum(study2_sample$gender == "Non-binary / third gender")`
non-binary / third gender;
`r sum(str_starts(study2_sample$gender, "Prefer") | is.na(study2_sample$gender))`
undisclosed gender; mean age = `r round(mean(study2_sample$age), 2)` years).
`r round(mean(study2_sample$chatgpt_used == "Yes", na.rm = TRUE) * 100, 0)`% of
these participants reported having used ChatGPT before (see 
@suppfig-chatgpt-study1).

### Design

We used a "control plus 2x2" between-subjects design. Participants were randomly 
allocated to either the control condition, in which people in the scenarios 
complete the tasks themselves, or one of four experimental conditions, in which 
people in the scenarios use AI to complete the tasks. In the experimental 
conditions, we manipulated whether people in the scenarios used AI as a 
collaborative tool or "fully" outsourced to AI, and whether people were honest 
or deceptive about their use of AI. This resulted in five conditions overall: 
(*i*) the control condition, (*ii*) the tool-honest condition, (*iii*) the 
tool-deception condition, (*iv*) the full-honest condition, and (*v*) the 
full-deception condition.

### Procedure

We presented participants with six scenarios. Each scenario described a person 
completing a task, such as writing computer code or writing a love letter. The 
six tasks were randomly drawn from a larger set of 16 tasks (see @supptbl-tasks 
for the full list of tasks). For each scenario, we first told participants:

- *Control condition*: "In order to complete this task, [the person] works on it 
by themselves from start to finish."
- *Tool outsourcing conditions*: "In order to complete this task, [the person] 
uses the AI tool ChatGPT. They ask ChatGPT to provide ideas, inspiration, and 
feedback, but they edit and rewrite the suggestions and finish the task
themselves."
- *Full outsourcing conditions*: "In order to complete this task, [the person]
uses the AI tool ChatGPT. They copy ChatGPT's output word-for-word, rather than
doing it themselves."

We then told participants in the experimental conditions:

- *Honest conditions*: "After completing the task, [the person] is asked how
they came up with their ideas. [The person] acknowledges that they used ChatGPT
as a tool / got ChatGPT to do the task for them."
- *Deception conditions*: "After completing the task, [the person] is asked how
they came up with their ideas. [The person] does not acknowledge that they used 
ChatGPT as a tool / got ChatGPT to do the task for them."

We then asked participants how well each of the following words described the 
person in the scenario: competent, warm, moral, lazy, and trustworthy. 
Participants answered these questions on 7-point Likert scales, ranging from 
"does not describe [the person] well" to "describes [the person] extremely 
well".

After the six scenarios, we asked participants several questions about the AI 
tool ChatGPT, including their familiarity with ChatGPT, whether they had used 
ChatGPT before, how frequently they used ChatGPT, and how trustworthy they 
thought ChatGPT was (see @suppfig-chatgpt-study1).

### Pre-registration

We pre-registered the study on the Open Science Framework
(<https://osf.io/knswr>).

### Statistical Analysis

We fitted Bayesian multivariate multilevel cumulative-link ordinal models to the
data using the *brms* R package [@Burkner2017]. We modelled each character 
evaluation -- competence, warmth, morality, laziness, and trustworthiness -- as 
a separate response variable and included fixed effects for conditions, varying 
intercepts for participants, and varying intercepts and slopes for tasks. We 
used regularising priors for all parameters to impose conservatism on parameter 
estimates (see Supplementary Materials for full model specifications). All 
models converged normally ($\hat{R}$ ≤ 1.01).

### Transparency and Openness

For all studies in this paper, we report how we determined our sample size, all
data exclusions, all manipulations, and all measures in the studies. All studies
were pre-registered. Analyses for all studies were conducted in R v4.4.2 
[@RCoreTeam]. Visualisations were produced using the *ggplot2* and *patchwork* 
packages [@Wickham2016; @Pedersen2025]. The manuscript was reproducibly 
generated using the *targets* package [@Landau2021] and *quarto* [@Allaire2024].
All data and code to reproduce the analyses and figures in this paper can be 
found here: <https://github.com/ScottClaessens/outsourcing>

## Results

Across all five character evaluations, we found that fully outsourcing to AI 
(i.e., copying the AI output verbatim) was perceived more negatively than using
AI as a collaborative tool (@fig-treatments-study1; 
@tbl-treatment-diffs-study1). By contrast, we found that deception about AI
usage had specific negative effects on perceptions of morality and
trustworthiness: people who did not acknowledge their use of AI were perceived
as less moral and less trustworthy. We did not find any interaction effects
between full outsourcing and deception.

```{r, fig.height=6, fig.width=6}
#| label: fig-treatments-study1
#| fig-cap: Character Evaluations in Study 1
#| fig-align: center
#| apa-note: Participants in the control condition and four AI outsourcing conditions evaluated people in the scenarios on (a) competence, (b) warmth, (c) morality, (d) laziness, and (e) trustworthiness. Coloured points represent participant responses to the questions, jittered for easier viewing. Black points are estimated marginal means from the fitted model, pooling over participants and tasks. Black points and line ranges represent posterior medians and 95% credible intervals, respectively.

tar_read(plot_treatments_study2)
```

```{r}
#| label: tbl-treatment-diffs-study1
#| tbl-cap: Pairwise Contrasts in Study 1 \vspace{20pt}
#| apa-note: Numbers reflect differences in marginal means on a 7-point Likert scale, pooling over participants and tasks. The bottom row represents the interaction between full outsourcing and deception (i.e., the difference between the differences in the rows above). Main numbers are posterior medians, numbers in the square brackets are 95% credible intervals.

tar_read(table_treatment_diffs_effects_study2) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    linesep = ""
    ) %>%
  kable_styling(
    font_size = 8,
    position = "left"
    ) %>%
  add_header_above(header = c(" " = 1, "Response" = 5)) %>%
  pack_rows(
    index = c(
      "Comparison to control" = 4,
      "Effect of full outsourcing" = 2,
      "Effect of deception" = 2,
      "Interaction effect"
      )
    ) %>%
  # fix overlapping apa note
  paste0("\\vspace{20pt}") %>%
  knitr::asis_output()
```

The effects of outsourcing to AI varied across the different tasks, especially 
for perceptions of warmth and morality (@fig-treatments-tasks-study1). For 
example, people who used AI for social tasks, such as writing an apology letter
or a bereavement card, were perceived as less warm, less moral, and lazier 
compared to people who completed the task themselves. This was true regardless
of whether AI was used as a tool or fully delegated to and whether the person
was honest or deceptive about their use of AI. By contrast, we observed weaker
effects of outsourcing for non-social tasks like writing computer code or 
planning a syllabus.

::: {.landscape}

```{r, fig.height=5, fig.width=8.5}
#| label: fig-treatments-tasks-study1
#| fig-cap: Variation in the Effects of Outsourcing across Tasks in Study 1 \vspace{-30pt}
#| fig-align: center
#| apa-note: Tasks are ordered from most social (top) to least social (bottom) according to ratings from a pilot study. Point ranges are differences in marginal means on a 7-point Likert scale for the honest conditions (red) and deception conditions (blue) compared to the control condition. Upper panels refer to the tool outsourcing conditions, and lower panels refer to the full outsourcing conditions. Points and ranges represent posterior medians and 95% credible intervals, respectively.

tar_read(plot_treatments_tasks_study2)
```

:::

To determine the factors that predict variation across tasks, we incorporated 
ratings of tasks from a pilot study (see Supplementary Materials for details).
Participants were asked to rate the different tasks on several features: whether
the task is social, requires social skills, impacts others, has important 
consequences, and requires effort. All of these features predicted stronger 
causal effects of outsourcing to AI compared to the control (Supplementary 
Figures [-@suppfig-interactions-study1] and 
[-@suppfig-interaction-pars-study1]). In other words, outsourcing to AI is 
perceived more negatively for tasks that have these features, compared to tasks
without these features.

## Discussion

In this study, we manipulated whether people used AI as a collaborative tool or 
copied the AI's output verbatim. In line with our predictions, we found that 
"fully" outsourcing to AI was perceived more negatively than using AI as a 
collaborative tool, particularly for socio-relational tasks. These negative 
perceptions of people who fully outsource their work to AI may be driven in part
by a general suspicion in unsupervised AI output due to what has been termed 
"algorithmic aversion" -- the tendency for people to distrust AI relative to 
humans [@Dietvorst2015]. This aversion is particularly strong in the moral 
domain [@Bigman2018]. Our results suggest that people may not only harbour 
negative perceptions of AIs themselves, but also of humans who unthinkingly copy
the output from these AIs, especially for socio-relational tasks.

We also manipulated whether people were honest or deceptive in their use of AI.
We found that people were seen as less moral and less trustworthy if they did
not acknowledge their use of AI when asked how they came up with their ideas.
This builds on previous work showing that people dislike and distrust those who
lie [@Tyler2006]. However, passing off AI-generated work as one's own is a 
specific kind of lie -- it involves taking undeserved credit for the work of a 
machine and misrepresenting one's own emotional capabilities and writing skills.
Interestingly, we found that failing to acknowledge use of AI had negative 
effects on perceptions of morality and trustworthiness across all the tasks we 
studied, suggesting that this phenomenon generalises across a range of 
socio-relational and professional tasks.

We have shown outsourcing to AI leads to negative character evaluations, 
especially when the AI output is copied verbatim and the work is passed off as 
one's own. However, it remains unclear whether these negative perceptions also
extend to the work itself. It could be, for example, that someone is perceived
badly for using ChatGPT to write their bereavement card, but the actual quality
of the writing is perceived to be just as good, if not better, than if the
person had written the card themselves. Indeed, evidence suggests that text
generated by ChatGPT is rated as higher quality than human-written text
[@Noy2023]. However, because this previous study used treatment-blind raters, it
was unable to assess whether knowing that a piece of text was AI-generated
affects how people perceive the text, while holding the quality of the text
constant. We explored this possibility in Study 2.

@@

## Methods

### Participants

We conducted a power simulation to determine our target sample size. The 
simulation suggested that a sample size of 125 participants per condition 
(overall *n* = 750 for six conditions) would be required to detect a 
small-to-medium difference between conditions (Cohen's *d* ≈ 0.40) with above 
80% power.

```{r}
study3_sample <-
  tar_read(study3_data) %>%
  group_by(id) %>%
  summarise(
    gender = unique(gender),
    age = unique(age),
    chatgpt_used = unique(chatgpt_used)
    )
```

We recruited a convenience sample of 800 participants from the United Kingdom 
through Prolific. After excluding participants who failed our pre-treatment 
attention check, we were left with a final sample of `r nrow(study3_sample)` 
participants (`r sum(study3_sample$gender == "Female")` female; 
`r sum(study3_sample$gender == "Male")` male;
`r sum(study3_sample$gender == "Non-binary / third gender")`
non-binary / third gender;
`r sum(str_starts(study3_sample$gender, "Prefer") | is.na(study3_sample$gender))`
undisclosed gender; mean age = `r round(mean(study3_sample$age), 2)` years).
`r round(mean(study3_sample$chatgpt_used == "Yes", na.rm = TRUE) * 100, 0)`% of
these participants reported having used ChatGPT before (see 
@suppfig-chatgpt-study2).

### Design

We randomly allocated participants to one of six conditions in a 3x2 
between-subjects design. We manipulated the type of outsourcing: (*i*) no 
outsourcing control, (*ii*) using AI as a tool, and (*iii*) fully outsourcing to
AI. We also manipulated whether the task prompt was social or non-social.

### Procedure

We told participants that they would read and evaluate a short piece of writing
from "another participant" (in reality, we generated the writing using ChatGPT 
version 4o and edited the text to appear more human-like). The prompt for the 
piece of writing varied between conditions:

- *Social conditions*: "Please write a description of a close family member or 
friend, explaining why they are special to you."
- *Non-social conditions*: "Please write a short description of a book, TV show,
or film of your choice."

We explained that the "other participant" was asked several questions about how
they produced their answer, including whether or not they used an AI tool like
ChatGPT. We explained that the participant was encouraged to be honest and told
that they would be paid regardless. The response from the "other participant" 
varied between conditions:

- *Control conditions*: "The participant reported that they did not use any AI 
tool like ChatGPT. Instead, they worked on the response themselves from start to
finish."
- *Tool outsourcing conditions*: "The participant reported using ChatGPT to 
provide ideas, inspiration, and feedback. The participant told us that they 
edited and rewrote ChatGPT's suggestions and finished writing the response 
themselves."
- *Full outsourcing conditions*: "The participant reported using ChatGPT to 
complete the task. The participant told us that they copied ChatGPT's output
word-for-word, rather than producing the response themselves."

We then presented participants with a randomly-chosen pre-generated essay answer
to the prompt (see Supplementary Tables [-@supptbl-essay-answers-social-study2] 
and [-@supptbl-essay-answers-nonsocial-study2] for full essay answers). In the 
social conditions, the answer either referred to the participant's father, their
sister, or their best friend. In the non-social conditions, the answer either
referred to the book The Hobbit, the TV show Buffy the Vampire Slayer, or the
film Titanic. Reading times and responses to a follow-up comprehension question
suggested that participants read the essay answers in sufficient detail (see 
@supptbl-essay-comprehension-study2).

We asked participants about their perceptions of the essay answer and the "other
participant". We asked how well-written, meaningful, and authentic they thought
the answer was (7-point Likert scales), what letter grade they would give the
answer (A-E), and how much they would hypothetically reward the other
participant for their work (from £0.00 to £1.00). We also asked how well each of
the following words described the other participant: competent, warm, moral,
lazy, and trustworthy (7-point Likert scales).

At the end of the study, we gave participants a manipulation check and asked
them whether they believed the manipulation. Responses to these questions
suggested that the outsourcing manipulation was successful (see 
@supptbl-manipulation-check-study2). We also asked participants several 
questions about ChatGPT (see @suppfig-chatgpt-study2).

### Pre-registration

We pre-registered the study on the Open Science Framework[^1].

[^1]: Due to a technical error with archiving this pre-registration on the Open
Science Framework, the timestamp for the registration was lost. However, on our
OSF project (<https://osf.io/xhmzk>), it is possible to view our 
pre-registration document file and its timestamped upload date.

### Statistical Analysis

We fitted two Bayesian multilevel models to the data. The first model was a 
multivariate cumulative-link ordinal model including all Likert scales as 
separate response variables. The second model was a zero-one-inflated-beta model
applied specifically to the reward question, which was a slider scale varying
between 0 and 1. For both models, we included fixed effects for the interaction
between outsourcing type and task type and varying intercepts and slopes for
essay answers. We used regularising priors for all parameters to impose 
conservatism on parameter estimates (see Supplementary Materials for full model 
specifications). All models converged normally ($\hat{R}$ ≤ 1.01).

## Results

We found that essay responses that were ostensibly generated using AI were 
perceived as less meaningful and less authentic, but just as well-written, 
compared to essay responses ostensibly written by a human
(@fig-treatments-work-study2; @tbl-treatment-diffs-work-study2). Participants 
also marked AI-generated essays with a lower grade and rewarded AI-generated 
essays with a lower hypothetical monetary bonus.

```{r, fig.height=5, fig.width=7}
#| label: fig-treatments-work-study2
#| fig-cap: Perceptions of the Work in Study 2
#| fig-align: center
#| apa-note: Participants in the control condition, the tool outsourcing condition, and the full outsourcing condition evaluated the essay response to the writing task. Participants rated whether the essay response was (a) well-written, (b) meaningful, and (c) authentic. Participants also (d) graded the work and (e) rewarded the work with a hypothetical monetary bonus. Jittered points represent participant responses to the questions, split by whether the writing task was a non-social task (red) or a social task (blue). Point ranges are estimated marginal means from the fitted model, pooling over essay answers. Points and line ranges represent posterior medians and 95% credible intervals, respectively.

tar_read(plot_treatments_work_study3)
```

```{r}
#| label: tbl-treatment-diffs-work-study2
#| tbl-cap: Pairwise Contrasts for Perceptions of the Work in Study 2 \vspace{20pt}
#| apa-note: Numbers reflect differences in marginal means on either a 7-point Likert scale (well-written, meaningful, authentic), a 5-point ordinal grade scale (grade), or a 0-1 sliding scale (reward). Estimates are pooled over essay answers. The bottom rows represent the interactions between outsourcing type and task type. Main numbers are posterior medians, numbers in the square brackets are 95% credible intervals.

tar_read(table_treatment_diffs_effects_study3) %>%
  dplyr::select(1:6) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    linesep = ""
  ) %>%
  kable_styling(
    font_size = 7.5,
    position = "left"
    ) %>%
  add_header_above(header = c(" " = 1, "Response" = 5)) %>%
  pack_rows(
    index = c(
      "Effect of outsourcing type" = 6,
      "Effect of task type" = 3,
      "Interaction effect" = 3
      )
    ) %>%
  pack_rows("\u00A0\u00A0Task type = Social", 1, 3, indent = FALSE) %>%
  pack_rows("\u00A0\u00A0Task type = Non-social", 4, 6, indent = FALSE) %>%
  # fix overlapping apa note
  paste0("\\vspace{20pt}") %>%
  knitr::asis_output()
```

We did not find differences in perceptions of the work between the tool 
outsourcing and full outsourcing conditions, except for the reward question,
where fully outsourced essays (i.e., essays copied verbatim from ChatGPT) were
rewarded £0.23 less than essays generated using AI as a collaborative tool. We
did not find any differences between social and non-social tasks and did not
find any interaction effects.

Turning to character evaluations, we found that people were perceived more
negatively across all character evaluations if they outsourced the writing task
to AI, either by using ChatGPT as a collaborative tool or by copying the text
from ChatGPT verbatim (@suppfig-treatments-person-study2; 
@supptbl-treatment-diffs-person-study2). However, we did not find any 
differences in character evaluations between the tool outsourcing and full 
outsourcing conditions. We did not find differences in character evaluations 
between social and non-social tasks and did not find any interaction effects.

## Discussion

Extending our previous results, we found that not only are people perceived 
negatively for outsourcing their work to AI, but the work produced by AI is 
itself also perceived negatively. In particular, text that has purportedly been
generated using AI is perceived to be less meaningful, less authentic, worthy of
a lower grade, and worthy of a lower monetary reward compared to human-generated
text. This is despite the actual text itself being held constant, perhaps
explaining why perceived writing quality did not consistently differ between
conditions.

Unlike in Study 1, we found no differences in the effect of AI-outsourcing
between social and non-social tasks. This may be due to the particular tasks we
chose. Writing *about* someone close to you is not quite the same as writing 
something *for* someone close to you, as is the case with wedding vows, love 
letters, and bereavement cards. We also found no differences between the tool
and full outsourcing conditions, aside from the lower rewards given to
participants in the latter condition. It is possible that differences between
conditions were smaller under this design and we were not sufficiently powered
to detect them.

Thus far, we have presented evidence that outsourcing tasks to AI causes
negative character evaluations and negative perceptions of the work. But the
mechanisms underlying these effects remain unclear. We have already suggested
that effort may play a role, since perceived effort (or lack thereof) is often
used as a signal of one's moral character [@Cubitt2011] and cooperative intent
[@Celniker2023]. Study 2 also suggested a role of authenticity: people who 
outsource to AI may be perceived as producing work that is less authentically
their own, leading to negative evaluations [see also @Kirk2025]. Indeed,
previous studies have shown that people react negatively to individuals and 
brands that behave inauthentically by strategically presenting themselves in
ways that are misaligned with their true nature [@Sedikides2024; @Silver2021].

To explore the potential mechanisms of effort and authenticity, we return to our
previous design and experimentally manipulate (1) how much effort people put
into the task and (2) whether people outsource the task to a standard large
language model like ChatGPT or a personalised model that has been trained
specifically on one's own prior writings. The latter authenticity manipulation
is inspired by recent psychological work on the credit-blame asymmetry in AI use
showing that people receive more personal credit for their work when they use an
AI model trained on their own prior writings [@Earp2024]. If effort and 
authenticity are important mechanisms underlying negative perceptions of 
outsourcing to AI, then these manipulations should influence character 
evaluations.

@@

## Methods

### Participants

```{r}
study4_sample <-
  tar_read(study4_data) %>%
  group_by(id) %>%
  summarise(
    gender = unique(gender),
    age = unique(age),
    chatgpt_used = unique(chatgpt_used),
    effort = unique(effort),
    authentic = unique(authentic)
    )
```

We used the same power estimate from Study 1 to determine our target sample size
of *n* = 750 (150 participants in each of five conditions). We recruited a 
convenience sample of 802 participants from the United Kingdom through Prolific.
After excluding participants who failed our pre-treatment attention check, we
were left with a final sample of `r nrow(study4_sample)` 
participants (`r sum(study4_sample$gender == "Female")` female; 
`r sum(study4_sample$gender == "Male")` male;
`r sum(study4_sample$gender == "Non-binary / third gender")`
non-binary / third gender;
`r sum(str_starts(study4_sample$gender, "Prefer") | is.na(study4_sample$gender))`
undisclosed gender; mean age = `r round(mean(study4_sample$age), 2)` years).
`r round(mean(study4_sample$chatgpt_used == "Yes", na.rm = TRUE) * 100, 0)`% of
these participants reported having used ChatGPT before (see 
@suppfig-chatgpt-study3).

### Design

We used the same "control plus 2x2" between-subjects design as in Study 1. In 
the experimental conditions, we manipulated whether people in the scenarios used
a standard or personalised AI model, and whether people put more or less effort
into the task. This resulted in five conditions overall: (*i*) the control 
condition, (*ii*) the standard-low-effort condition, (*iii*) the 
standard-high-effort condition, (*iv*) the personalised-low-effort condition, 
and (*v*) the personalised-high-effort condition.

### Procedure

The procedure was mostly identical to Study 1, but we updated the study preamble
and the presentation of the scenarios. For participants in the personalised AI
conditions, we expanded the study preamble to explain that personalised AI
models were trained on people's own prior writings and "tailored to each 
specific person and their own thoughts, feelings, and values". Then in the 
scenarios, we told participants in the experimental conditions:

- *Standard AI conditions*: "In order to complete this task, [the person] uses 
the AI tool ChatGPT."
- *Personalised AI conditions*: "In order to complete this task, [the person]
uses a personalised AI tool."

We then told participants:

- *Low effort conditions*: "[The person] quickly gives the AI a rushed prompt 
and uses its first output."
- *High effort conditions*: "[The person] carefully gives the AI several 
detailed prompts and, after multiple rounds of changes, uses its resulting 
output."

At the end of the study, we asked participants to choose which of these was more
authentic and effortful, respectively. 
`r round(mean(str_detect(study4_sample$authentic, fixed("personalised")), na.rm = TRUE) * 100)`%
of participants stated that the personalised AI was more authentic and 
`r round(mean(str_detect(study4_sample$effort, fixed("Carefully"))) * 100)`% of 
participants stated that giving the AI several detailed prompts was more 
effortful.

### Pre-registration

We pre-registered the study on the Open Science Framework
(<https://osf.io/vaq7u>).

### Statistical Analysis

We fitted the same Bayesian multivariate multilevel cumulative-link ordinal 
models as in Studies 1 and 2. All models converged normally ($\hat{R}$ ≤ 1.01).

## Results

We found that people who outsourced to AI in a low effort way were perceived as
less competent, less moral, lazier, and less trustworthy than people who put
more effort into their use of AI (@fig-treatments-study3;
@tbl-treatment-diffs-study3). By contrast, we found that character evaluations
did not differ between people who used a standard AI model rather than a 
personalised AI model. We also found no interaction effects between effort and 
the type of AI used.

```{r, fig.height=6, fig.width=7}
#| label: fig-treatments-study3
#| fig-cap: Character Evaluations in Study 3
#| fig-align: center
#| apa-note: Participants in the control condition and four AI outsourcing conditions evaluated people in the scenarios on (a) competence, (b) warmth, (c) morality, (d) laziness, and (e) trustworthiness. Coloured points represent participant responses to the questions, jittered for easier viewing. Black points are estimated marginal means from the fitted model, pooling over participants and tasks. Black points and line ranges represent posterior medians and 95% credible intervals, respectively.

tar_read(plot_treatments_study4)
```

```{r}
#| label: tbl-treatment-diffs-study3
#| tbl-cap: Pairwise Contrasts in Study 3 \vspace{20pt}
#| apa-note: Numbers reflect differences in marginal means on a 7-point Likert scale, pooling over participants and tasks. The bottom row represents the interaction between AI type and effort (i.e., the difference between the differences in the rows above). Main numbers are posterior medians, numbers in the square brackets are 95% credible intervals.

tar_read(table_treatment_diffs_effects_study4) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    linesep = ""
    ) %>%
  kable_styling(
    font_size = 7,
    position = "left"
    ) %>%
  add_header_above(header = c(" " = 1, "Response" = 5)) %>%
  pack_rows(
    index = c(
      "Comparison to control" = 4,
      "Effect of AI type" = 2,
      "Effect of effort" = 2,
      "Interaction effect"
      )
    ) %>%
  # fix overlapping apa note
  paste0("\\vspace{20pt}") %>%
  knitr::asis_output()
```

As in Study 1, the effects of outsourcing to AI varied across the different 
tasks, especially for perceptions of warmth and morality 
(@suppfig-treatments-tasks-study3). We again found that the negative causal 
effects of outsourcing to AI were particularly strong for tasks that are social,
require social skills, impact others, have important consequences, and require
effort (Supplementary Figures [-@suppfig-interactions-study3] and 
[-@suppfig-interaction-pars-study3]).

## Discussion

We found that effort is an important mechanism by which outsourcing to AI leads
to negative character evaluations. People who engaged in effortless copying of
the AI's first output were perceived more negatively, whereas people who put in 
effort with multiple detailed prompts were perceived no differently from those 
who completed the tasks themselves. This result is in line with previous work
showing that effort is a signal of one's moral character [@Cubitt2011; 
@Celniker2023].
